---
title: "Full Information Maximum Likelihood"
author: "Dr. Iris Eekhout"
subtitle: "Workshop name"
date: "`r format(Sys.Date())`"
output: 
  ioslides_presentation:
    logo: Puzzle_zgnb3_m.jpg
    theme: readable
    css: eigencssfile.css
widescreen: true
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(mice)
library(lavaan)
library(dplyr)
library(semTools)
```

# Full Information Maximum Likelihood {data-background=Puzzle_zgnb3.jpg data-background-size=cover}

## How does it work

* Uses all observed data, so also partly observed rows, to estimate model parameters.
* Analysis and dealing with missing data, at once
* Different ways to estimate, for example EM algorithm
* Only involves variables used in the analysis


## FIML in `lavaan` {.smaller}

* In R use `lavaan` package with `missing = "fiml"`.

```{r echo=TRUE}
model <- '
  #variance
  Ozone ~~ Ozone
  Solar.R ~~ Solar.R
  Wind ~~ Wind
  Temp ~~ Temp
  
  #correlation
  Ozone ~~ Solar.R + Wind + Temp
  Solar.R ~~ Wind + Temp
  Wind ~~ Temp
  '
fit <- sem(model, data = airquality, missing = "fiml", meanstructure = TRUE)
```

## FIML descriptives {.smaller}


```{r echo=FALSE}
summary(fit, standardized = TRUE, header = FALSE)
```


## Output explained {.smaller}

Output is quite large and gives a lot of information at once. 

* Intercepts: FIML means 
  + *~1* in output tables 
* Variance: FIML variances 
  + *variable x ~~ variable x* in output tables
* Covariances: FIML correlation (because `standardized = TRUE`) 
  + *variable x ~~ variable y* in output tables


Use `fmi = TRUE` in the `summary()` function to get fraction of missing information. 

fmi = the relative increase in variance and decrease of precision due to missing data, i.e. impact of missing data on estimates.

## Missing data patterns 

* Missing data patterns in the data

```{r}

inspect(fit, 'patterns') 

```


## Missing data proportions

* A symmetric matrix where each element contains the proportion of observed datapoints for the corresponding pair of observed variables.

```{r}
inspect(fit, 'coverage')
```

## Linear regression analysis {.smaller}

```{r}
model <- '
  Ozone ~ Solar.R 
  '
fit_fiml <- sem(model, data = airquality, missing = "fiml")
summary(fit_fiml, header = FALSE, fmi = TRUE)
```

## Compare with MI {.smaller}

```{r}
imp_pmm <- mice(airquality %>% select(Ozone, Solar.R), method = "pmm", print = F)
fit_mi <- with(imp_pmm, lm(Ozone ~ Solar.R))
combi <- pool(fit_mi)
summary(combi)
```

## SEM with auxiliary variables {.smaller}

* Use auxiliary variables to improve missing data handling
* Auxiliary variables to covariances in the model

```{r}
model_aux <- '
  Ozone ~ Solar.R 
  Solar.R ~~ Wind + Temp + Month + Day
  Ozone ~~ Wind + Temp + Month + Day
  Wind ~~ Temp + Month + Day
  Temp ~~ Month + Day
  Month ~~ Day
  '
auxfit1 <- sem(model = model_aux, missing = "fiml", data = airquality)
```

## SEM with auxiliary variables {.smaller}

```{r, echo = FALSE}
summary(auxfit1, header = FALSE, fmi = TRUE)
```



## SEM with auxiliary variables {.smaller}

* Easily add auxiliary variables with `semTools` package

```{r}
aux.vars <- c("Wind", "Temp", "Month","Day")
auxfit <- sem.auxiliary(model=model, missing = "fiml", aux=aux.vars, fixed.x=FALSE, data=airquality)

```


## SEM with auxiliary variables {.smaller}

```{r echo = FALSE}
summary(auxfit, header = FALSE, fmi = TRUE)

```


## Compare with MI {.smaller}

```{r}
imp_pmm <- mice(airquality, method = "pmm", print = F)
fit_mi <- with(imp_pmm, lm(Ozone ~ Solar.R))
combi <- pool(fit_mi)
summary(combi)          
  
```
