---
title: "Software demo"
author: "Dr. Iris Eekhout"
subtitle: "Workshop name"
date: "`r format(Sys.Date())`"
output: 
  ioslides_presentation:
  logo: Puzzle_zgnb3_m.jpg
theme: readable
css: eigencssfile.css
widescreen: true
always_allow_html: yes
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(dmo)
library(mice)
library(haven)
```

# Software demo `mice` {data-background=Puzzle_zgnb3.jpg data-background-size=cover}


## Summary of data

```{r}
summary(nhanes2)

``` 


## Multiple imputation

```{r eval = FALSE, include = TRUE}
install.packages("mice")
library(mice)
```

```{r}
imp <-mice(nhanes2, m=5, maxit=10, seed = 800)
```

## predictor matrix

```{r}
imp$pred
```

Remove a predictor variable, e.g remove "bmi" as predictor for "chl".

```{r}

imp$pred["chl","bmi"] <- 0
imp$pred

pred_update <- imp$pred
```

## methode per variable

```{r}
imp$meth
```

Change the imputation method for one variable
```{r}
imp$meth["bmi"] <- "norm"
imp$meth

meth_update <- imp$meth
```


## Run with new settings

```{r}
imp_update <- mice(boys, m=5, maxit=10, pred=pred_update, meth=meth_update, seed = 90)

```


## Iteration plots

```{r}
plot(imp)
```

## View imputed values

```{r}
stripplot(imp)
```

## View imputed values linear

```{r}
xyplot(imp, chl ~ bmi)

```

## Analyze imputed data

```{R}
fit <- with(imp, lm(chl ~ age))

```


## Pool results

```{r}

pool(fit)

```


