---
title: "Multilevel imputation"
author: "Dr. Iris Eekhout"
subtitle: "Workshop name"
date: "`r format(Sys.Date())`"
output: 
  ioslides_presentation:
    logo: Puzzle_zgnb3_m.jpg
    theme: readable
    css: eigencssfile.css
widescreen: true
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nlme)
library(dplyr)
library(mice)
library(tidyr)
library(ggplot2)

theme_set(theme_light())
```

## Multilevel design

* Observations are (hierarchically) clustered in groups

* Example of a two-level structure
+ Student
+ School

* Example of three-level structure
+ Student
+ Class
+ School

## Other examples of multilevel studies

* Longitudinal data: Measurements in subjects
* Meta-analysis: data in studies
* Clinical studies: patients in hospitals

## Example data

* The `bdf` data set from the `nlme` package are language scores from grade 8 pupils in elementary schools in the Netherlands. 
* Pupils in classes in schools
* Variables at the lowest level (pupils)
* Variables at the higher levels (school-level or class-level)

## Data structure

```{R}
bdf %>% select(schoolNR, classNR, pupilNR, schoolSES, ses, IQ.verb, IQ.perf, langPOST ) %>% head(10)
```


## Multilevel data analysis

* Advanced regression analysis technique to take the correlations within the groups (or clusters) into account.
* Multilevel model
* Random effects model
* Mixed effect model
* Hierarchical regression model


## Random intercept

* Estimate of the variance of the intercepts for different schools (covariance adjustment/confounding)
* $Y_{ij} = \beta_0 + \beta_1*X + \beta_2*school + \epsilon_{ij}$
* $Y = \beta_0 + \beta_1*X + b_{0j} + \epsilon_{ij}$
+ $b_{0j}$ is random intercept at school level

```{r}
dat <- data.frame(x = seq(0:100), y = seq(0:100))
ggplot(dat, aes(x , y))+
  geom_blank() + 
  geom_abline(intercept = (2), slope = 0.3, color = 2) +
  geom_abline(intercept = (5), slope = 0.3, color = 3) +
  geom_abline(intercept = (8), slope = 0.3, color = 4) +
  geom_abline(intercept = (11), slope = 0.3, color = 5) +
  geom_abline(intercept = (14), slope = 0.3, color = 6) +
  geom_abline(intercept = (17), slope = 0.3, color = 7) 

```

## Random slope

* Estimate of the variance of the slopes for the different schools (interaction or effect-modification)
* $Y_{ij} = \beta_0 + \beta_1*X + \beta_2*school + \beta_3 * X * school + \epsilon_{ij}$
* $Y = \beta_0 + \beta_1*X + b_{1j} * X + \epsilon_{ij}$
+ $b_{1j}$ is random slope at school level

```{r}
dat <- data.frame(x = seq(0:100), y = seq(0:100))
ggplot(dat, aes(x , y))+
  geom_blank() + 
  geom_abline(intercept = (2), slope = 0.3, color = 2) +
  geom_abline(intercept = (2), slope = 0.4, color = 3) +
  geom_abline(intercept = (2), slope = 0.5, color = 4) +
  geom_abline(intercept = (2), slope = 0.6, color = 5) +
  geom_abline(intercept = (2), slope = 0.7, color = 6) +
  geom_abline(intercept = (2), slope = 0.8, color = 7) 

```

## Random intercept and slope

* $Y_{ij} = \beta_0 + \beta_1*X + \beta_2*school + \beta_3 * X * school + \epsilon_{ij}$
* $Y = \beta_0 + \beta_1*X + b_{0j} + b_{1j} * X + \epsilon_{ij}$
+ $b_{0j}$ is random intercept at school level
+ $b_{1j}$ is random slope at school level

```{r}
dat <- data.frame(x = seq(0:100), y = seq(0:100))
ggplot(dat, aes(x , y))+
  geom_blank() + 
  geom_abline(intercept = (2), slope = 0.3, color = 2) +
  geom_abline(intercept = (5), slope = 0.4, color = 3) +
  geom_abline(intercept = (8), slope = 0.5, color = 4) +
  geom_abline(intercept = (11), slope = 0.6, color = 5) +
  geom_abline(intercept = (14), slope = 0.7, color = 6) +
  geom_abline(intercept = (17), slope = 0.8, color = 7) 

```

## Missing data in multilevel data

* At measurement level (Pupil)
+ Outcome (e.g. language score)
+ Predictor (e.g. verbal IQ)
* At the higher level (class or school)
+ Predictor (e.g. school SES)
+ Class variable itself (e.g. school number)

## Missing data methods

* Missing data in the dependent variable only: **solved by maximum likelihood estimation of the model**
+ Estimation method uses all available data to estimate the model parameters that fit best with the data.

* Missing data in independent variable: multiple imputation
+ Imputation model should take the analysis model into account
+ Multilevel imputation

## Multilevel imputation

```{R}
bdfmiss <- bdf %>% mutate(prop = runif(n()),
                          IQ.verb = ifelse(prop < 0.25, NA, IQ.verb)) %>%
  select(schoolNR, classNR, pupilNR, schoolSES, ses, IQ.verb, IQ.perf, langPOST )
summary(bdfmiss)
## brandsma example?
  
```


